Efficient Computation of Extended Surface Sources

### 2022-10-25 repro/papers/GEM22/submitR1/paper.tex
mods

Source extension is one approach to avoiding this``cycle-skipping''
obstacle. It consists in imposing the wave equation as a soft
constraint, allowing the source field $\bff$ to have more degrees of
freedom than is permitted by a faithful model model of the seismic
experiment, and constraining these additional degrees of freedom by
means of an additive quadratic penalty modifying the problem
\ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } {\bf c}, \bff \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
\end{equation}
The linear operator $A$ penalizes deviation from known (or assumed)
characteristics of the source function - its null space consists of
feasible (or ``physical'') source models.
\cite{HuangNammourSymesDollizal:SEG19} present an overview of the
literature on source extension methods, describing a variety of
methods to add degrees of freedom to physical source model.

The present paper concerns {\em surface source extension}: physical
sources are presumed to be concentrated spatially at points $\bx_s$
in space, whereas their extended counterparts are permitted to spread
energy over a {\em source surface} containing the physical source spatial
locations. Similarly, the receiver spatial locations are confined to a
{\em receiver surface}. A simple choice for the penalty operator $A$ is then
multiplication by the distance $|\bx-\bx_s|$ to the physical source
location:
\begin{equation}
  \label{eqn:penop}
  (A\bff)(\bx,t) = |\bx-\bx_s|\bff(\bx,t)
\end{equation}
I shall use this choice of penalty operator whenever a specific choice
is necessary in the development of the theory below.

This paper presents a numerically efficient approach to solving the
{\em source subproblem} of problem \ref{eqn:esi}:
\begin{equation}
\label{eqn:esis}
\mbox{given } {\bf c}, \mbox{ choose } \bff \mbox{ to minimize }
\|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha \|A\bff\|^2 
\end{equation}
Solution of this subproblem is an essential component of {\em variable
  projection} algorithms for solution of the nonlinear inverse problem
\ref{eqn:esi}. Variable projection is not merely a convenient choice
of algorithm for this purpose: it is in some sense essential, see for
example \cite{Symes:SEG20}. It replaces the nonlinear
least squares problem \ref{eqn:esi} with a {\em reduced} problem, to
be solved iteratively. Each iteration involves solution of the
subproblem \ref{eqn:esis}. Therefore efficient solution of the
subproblem is essential to efficient solution of the nonlinear problem
via variable projection.

@@

I will use two 2D numerical models 
throughout to illustrate the theory. In both, horizontal lines serve as
source and receiver surfaces. The first is a ``crosswell'' or slab
configuration with an acoustic lens positioned between a deeps source
and a shallower line of 
receivers, resulting in markedly triplicated arrivals. The goal in
this first example is to construct a surface source that explains the data in a
homogeneous medium (that is, inversion in a wrong velocity, emulating
the early iterations of FWI based on extended modeling). The second is a layered
model in which velocity increases with depth, resulting the formation
of diving waves. This configuration simulates an
ocean-bottom node and a line of near-surface sources: the roles of
source and receiver are switched for computational convenience. In
this second example, the diving wave arrivals are isolated (by an
appropriate mute operator) and a
source constructed that explains them alone.

The next section defines the modeling operator $PL[{\bf c}]$, its
adjoint, and important specializations (pressure vs. normal velocity
sources and data). It also introduces the two 2D examples mentioned
above.  The following section constructs an approximate inverse of the
modeling operator by time reversal (as suggested by work in
photoacoustic tomography), and illustrates its efficacy in the two
examples. This construction requires extraction of
velocity data, or equivalently a surface source, from pressure data.
The subsequent sections describe this pressure-to-source operator,
express the approximate inverse as the modeling operator adjoint in
weighted norms (thus establishing that the modeling operator is {\em
  approximately unitary} in the sense of these norms), explain how to
use this construction to precondition Conjugate Gradient iteration,
and organize the preconditioning computation so as to involve only one
extra and relatively inexpensive wave propagation calculation. I use
the 2D examples to illustrate each of
these developments. The paper ends with a brief
discussion-and-conclusion section, reviewing what has been
accomplished and listing a few of the many questions left open.

@@

The preconditioner construction is closely related to the time
reversal method in photoacoustic tomography
\cite[]{StefanovUhlmannIP:09,Hristova:09}. A simplified mathematical
translation of this medical imaging task is to infer the initial
excess pressure distribution over a fluid-containing region at time
$t=0$ from measurements of the pressure on a surface enclosing the
fluid over a time interval $0 \le t \le t_{\rm max}$. This problem
clearly has strong similarities to, but also differences from, the
problem studied here. The time reversal method presumes that the
pressure field has returned to equilibrium (zero excess pressure), or close to it, at the
final time $t=t_{\rm max}$ Then the field can be (at least
approximately) viewed as the backwards-in-time initial boundary value
problem with zero final conditions at $t=t_{\rm max}$, and boundary
values given by the measurements. Evolving the field backwards in time
backwards to $t=0$ thus solves the problem. Except in special
circumstances, the pressure field never actually vanishes at finite
time, so the solution is approximate.

The seismic surface source extension proplem \ref{eqn:esis} differs in
several obvious ways from the photoacoustic setting. The measurement
or receiver surface does not surround the region of wave
propagation. It is not the initial pressure time-slice at $t=0$ that
is to be determined, but a time-extended source $\bff$ confined to a
surface. The penalty operator has no analogue in the basic
photoacoustic problem description. Finally, once these obstacles are
overcome, using the approximate solution operator to accelerate Krylov
iteration for solution of the optimization problem \ref{eqn:esis}
requires that the operator be identified as the adjoint of the
modeling operator with respect to suitable inner products in its domain
and range. In the pages to follow, I will address each of these issues.

First, reverse-time propagation can be localized via ray theory,
within high-frequency asymptotic approximation. This step requires
some assumptions about the ray fields: all rays responsible for significant
energy in the receiver data must arrive from the source surface, and must cross the source and receiver
surfaces transversally, that is, making non-zero angles. Also, the
rays must approach source and receiver surfaces from one side or the
other (``the inside''), so that locally the surfaces can be treated as boundaries of
propagation domains. With these assumptions, reverse-time propagation
of the receiver data
closely approximates the acoustic fields near the source surface. Next
observe that for a causal field with a surface pressure source (with a
continuous pressure field across the surface), the source is
proportional to the jump in the (particle) velocity field. Moreover,
to leading order in frequency, the velocity field switches sign at the
surface, so the jump is just twice the sampled value on the
surface. Thus reverse-time propagation and reading off the velocity
field on the source surface yields
a source that reproduces the pressure data, within an asymptotically small error.
%Preconditioning amounts to a change of
%norm in the domain and range spaces of the modeling operator. In this
%case, the modfied norms are weighted $L^2$, and the weight operators
%map pressure to corresponding surface source on the source and
%receiver planes.

This is the essence of the approximate solution of the problem
\ref{eqn:esis} for penalty weight $\alpha=0$. To see how Krylov
iteration might be accelerated, note that time-reverse propagation of
the receiver pressure values may be represented by a time-reverse
source propagation, with the source constructed from the velocity
field at the receiver surface. Moreover, the reverse-time
source-to-pressure propagation is the transpose (adjoint) of the
forward-time source-to-pressure propagation - this is a very simple
version of the adjoint state construction \cite[]{Plessix:06}. So the
approximate inverse is: conversion of pressure to source at the
receiver surface, followed by the adjoint of the modeling operator,
followed by the conversion of pressure to source at the source
surface. This sequence precisely describes the adjoint of the modeling
operator with respect to weighted norms in the spaces of source and
receiver data, with the weight operators being the
pressure-to-source operators. That this construction results in an
approximate inverse means that the modeling operator is approximately
unitary with respect to these weighted norms, which in turn implies
that the preconditioned conjugate gradient algorithm should converge rapidly.

@@

For acoustic wave physics, the coefficient vector is
$\bf{c}=(\kappa,\rho)^T$, with components bulk modulus $\kappa$ and
density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of
pressure $p$ (a scalar space-time field) and particle velocity $\bv$
(a vector space-time field). The wave operator $L[\bf{c}]$ is:
\begin{equation}
\label{eqn:aweop}
L[\bf{c}]\bf{u} = 
\left(
\begin{array}{c}
\frac{1}{\kappa}\frac{\partial p}{\partial t}  + \nabla \cdot \bv, \\
\rho\frac{\partial \bv}{\partial t} + \nabla p.
\end{array}
\right) 
\end{equation}
That is,
\begin{equation}
  \label{eqn:awemat}
  L[{\bf c}] = \left(
    \begin{array}{cc}
      \frac{1}{\kappa}\frac{\partial}{\partial t} & \nabla \cdot \\
      \nabla & \rho \frac{\partial}{\partial t}
    \end{array}
  \right)
\end{equation}
$L[{\bf c}]$ has a well-defined inverse in the sense of distributions
if it is restricted to either causal or anti-causal vector wavefields.

@@

Most of what follows is valid for any space dimension $n >0$. I will
describe the theory for $n=3$, write $\bx=(x,y,z)^T$ for the spatial
coordinate vector, and refer to the third (vertical) coordinate of
particle velocity $\bv$ as $v_z$. For computational convenience, the
examples are two-dimensional.

The surface source extension replaces point sources on or near a
surface in $\bR^3$ with source functions confined to the 
surface. The simplest example of this extended geometry specifies a
plane $\{(x,y,z,t): z=z_s\}$ at source depth $z_s$ as the surface. For
acoustic modeling, surface sources are combinations of constitutive law
defects and loads normal to the surface, localized on $z=z_s$. That
is, right-hand sides in the system $L\bu=\bff$ take the form
$\bff(\bx,t) = (h_s(x,y,t)\delta(z-z_s),
f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar defect $h_s$ and normal
force $f_s$ ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in
\ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$
takes the form
\begin{eqnarray}
\label{eqn:awepm}
\frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} & = & - \nabla \cdot \bv^{\pm} +
h_s \delta(z-z_s), \nonumber \\
\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} +
                                                f_s{\bf e} \delta(z-z_s),\nonumber \\
p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
\bv^{\pm} & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}

\noindent {\bf Remark:} In system \ref{eqn:awepm} and many similar
systems to follow, I will use the shorthand
\[
  p^+ = 0 \mbox{ for } t \ll 0 
\]
to mean that $p^+$ is {\em causal}, that is,
\[
  \mbox{For some } T \in \bR, p^+(\cdot,t) = 0 \mbox{ for all } t <
  T.
\]
Similarly,
\[
  p^- = 0 \mbox{ for } t \gg 0 
\]
signifies that $p^-$ is anti-causal.

Extended forward modeling consists in solving \ref{eqn:awepm} and
sampling the solution components at receiver locations. For
simplicity, throughout this paper I will assume that the receivers are
located on another spatial plane $\{(x,y,z,t): z=z_r\}$ at receiver
depth $z_r$. The constructions to follow involve interchange of the
roles of $z_s$ and $z_r$ (that is, locating sources on $z=z_r$ and
receivers at $z=z_s$), so both require sampling operators: $P_s,P_r$
are the sampling operators on $z=z_s$, $z=z_r$ respectively. In
practice, sampling necessarily occurs at a discrete array of points
(trace locations), and over a zone of finite extent. In this
theoretical discussion, I will neglect the finite sample rate, and
regard the data, for example $P_rp^+$, as continuously sampled. The
output samples are necessarily muted, that is, non-zero only over a space-time domain
of finite extent. This mute, and any tapering applied to the data
traces, are regarded as part of the sampling operators $P_s,P_r$.

The causal/anti-causal vector
modeling operators ${\cal S}^{\pm}_{z_s,z_r}$ are defined in terms of
the solutions $(p^{\pm},\bv^{\pm})$ of the systems \ref{eqn:awepm} by
\begin{equation}
  {\cal S}^{\pm}_{z_s,z_r}(h_s,f_s)^T  = (P_rp^{\pm},P_r v_z^{\pm})^T,
  \label{eqn:fwd}
\end{equation}
The subscript signifies that sources are located on $z=z_s$,
receivers on $z=z_r$. It is necessary to include this information in
the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
several locations will be needed later in the discussion.

Denote by $\Pi_i, i=0,1$ the projection on the first,
respectively second, component of a vector in $\bR^2$. The 
forward modeling operator from pressure source to pressure trace is
\begin{equation}
  \label{eqn:sdef}
  S^{\pm}_{z_s,z_r} = \Pi_0 {\cal S}^{\pm}_{z_s,z_r} \Pi_0^T 
\end{equation}
and the forward modeling operator from velocity source (normal force)
to velocity trace is
\begin{equation}
  \label{eqn:vdef}
  V^{\pm}_{z_s,z_r} = \Pi_1 {\cal S}^{\pm}_{z_s,z_r} \Pi_1^T 
\end{equation}

With these conventions, we can write the version of the source
subproblem \ref{eqn:esis} studied in this paper as
\begin{equation}
  \label{eqn:esisp}
  \mbox{find }h_s\mbox{ to minimize }\|S^{+}_{z_s,z_r}h_s- d\|^2 +
  \alpha^2\|Ah_s\|^2.
\end{equation}

The two examples mentioned in the introduction illustrate the setting
just described. The first example (Figure \ref{fig:olensrays0}) embeds
an acoustic (low-velocity) lens between a source surface at $z_s=3000$
(orange horizontal line) 
m and a receiver surface at $z_r=1000$ m (yellow horizontal line).  The data to be used in this
example results from a point source at $z_s=3500, x_s=3500$ m (orange
star). I have overlain the rays connecting this source point with the
data portion of the receiver surface: evidently all of the rays
involved cross the source and receiver surfaces transversally. Also,
regarding ``inwards'' as being ``upwards'' at the source surface,
``downwards'' at the receiver surface, it is clear that the data is
incoming at the source surface, outgoing at the receiver
surface.

@@

%\ref{fig:drecplh0}, \ref{fig:dfwdplh0}
Figure \ref{fig:drecplh0} shows the gather $P_rp_{\rm pt}^{+}$, where
$p_{\rm pt}^{+}$ is causal field generated by the point source at
$(3500,3500)$ m with a trapezoidal bandpass filter wavelet having
significant energy between 1 and 12.5 Hz. The mute embedded in $P_r$
limits trace positions to 2000 m $ \le x_r \le $ 6000 m, and times to
1.2 s $ \le t \le $ 3 s.

Figure \ref{fig:dsrcphh0} displays an extended (pressure) source $h_s$, in the form of
traces in the spatial range 2000 m $ \le x_r \le $ 6000 m and time
range 0 s $\le t \le $ 2 s. The modeling operator output $S_{z_s,z_r}^+h_s$,
obtained from the causal solution $(p^+, \bv^+)$ of \ref{eqn:awepm}
with this choice of $h_s$ and $f_s=0,$ by application of the sampling
operator $P_r$ to $p^+$, are shown in Figure \ref{fig:dfwdplh0}.

\noindent {\bf Remark:} The close resemblance between Figures
\ref{fig:drecplh0} and \ref{fig:dfwdplh0} is not accidental  - the
extended source $h_ s$ shown in Figure \ref{fig:dsrcphh0} is constructed
so that $S^++_{z_s,z_r}h_s$ (Figure \ref{fig:dfwdplh0}) closely
approximates the point source gather $P_r p^+_{\rm pt}$ (Figure
\ref{fig:drecplh0}). This construction will be explained in the next section.

@@

This geometry permits inversion of the diving wave data alone, which
is isolated via a mute, depicted in (Figure \ref{fig:ndwmute0}. The
mute includes time truncation at 6 s, and offsets between -10000 and
10000 m. The isolated
diving wave (window between offsets 4000 and 10000) is shown in
Figure \ref{fig:ndwdata0}.

\begin{figure}
  \centering
  \subfloat[\label{fig:ndwraw0}]{\includegraphics[width=0.32\textwidth]{ndwraw0.pdf}}
  \subfloat[\label{fig:ndwmute0}]{\includegraphics[width=0.32\textwidth]{ndwmute0.pdf}}
  \subfloat[\label{fig:ndwdata0}]{\includegraphics[width=0.32\textwidth]{ndwdata0.pdf}}
  \caption{Data generated using configuration in Figure
    \ref{fig:ooplrays0}. (a) traces from point source at $(500,10000)$ m
    with $[1, 2, 7.5, 12.5]$ Hz zero phase trapezoidal bandpass
    filter wavelet, delayed by $0.4$ s. (b) Mute function,
    incorporated in the sampling operator $P_r$. (c) Diving wave
    gather. Color scale is same for (a) and (c).}
\end{figure}
%As explained in the Introduction, the downgoing constraint on the square-integrable
%source functions $h_s, f_s$ is
%essential both for finite energy solutions of the system
%\ref{eqn:awepm} to exist, and for these solutions to have well-defined
%traces on the receiver surface $z=z_r$. This constraint will be
%assumed throughout, often tacitly.
%For downgoing solutions of system \ref{eqn:awepm}, the key
%components ($p^{\pm}$ and $v^{\pm}_z$) are continuous functions of $z$
%in the open slab $z_s<z<z_r$ with well-defined limits at the boundary
%planes, but may be discontinuous at the source plane
%$z=z_s$. Similarly, the roles of $z_s$ and $z_r$ will be interchanged
%in some of the constructions to come, and the corresponding solutions
%may be discontinuous at $z=z_r$. Accordingly, interpret $P_s$, $P_r$
%as the limit from right and left respectively: for $u=p^{\pm}$ or
%$v^{\pm}_z$,
%\begin{eqnarray}
 % \label{eqn:defsamp}
 % P_su(x,y,t) &=& \lim_{z \rightarrow z_s^+} u(x,y,z,t),\nonumber \\
 % P_ru(x,y,t) &=& \lim_{z \rightarrow z_r^-} u(x,y,z,t).                  
%\end{eqnarray}

>>>>

Source extension is one approach to avoiding this``cycle-skipping''
obstacle. It consists in imposing the wave equation as a soft
constraint, allowing the source field $\bff$ to have more degrees of
freedom than is permitted by a faithful model model of the seismic
experiment, and constraining these additional degrees of freedom by
means of an additive quadratic penalty modifying the problem
\ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } {\bf c}, \bff \mbox{ to minimize } \|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
\end{equation}
The linear operator $A$ penalizes deviation from known (or assumed)
characteristics of the source function - its null space consists of
feasible (or ``physical'') source models.

Well-studied examples of extended source approaches to FWI include
Wavefield Reconstruction Inversion (WRI)  \cite[]{LeeuwenHerrmannWRI:13,LeeuwenHerrmann:16,Lietal:18,Aghmiryetal:20,Louboutinetal:20} and Adaptive Waveform
Inversion (AWI)
\cite[]{Warner:16,GuachWarnerRavaut:GEO19,Guaschetal:NPJDM20,Yongetal:EAGE21,Warneretal:SEG21}. \cite{HuangNammourSymesDollizal:SEG19}
present an overview of the recent
literature on source extension methods.

The present paper concerns {\em surface source extension}: physical
sources are presumed to be concentrated at points in space, whereas
their extended counterparts are permitted to spread energy over a {\em
  source surface}. Similarly, receivers are located on a {\em receiver
  surface}. Assuming that the presumed point source is at location
$\bx_s$, a simple (though certainluy not the only) choice for the
penalty operator $A$ is then multiplication by the distance
$|\bx-\bx_s|$ to the physical source location:
\begin{equation}
  \label{eqn:penop}
  (A\bff)(\bx,t) = |\bx-\bx_s|\bff(\bx,t)
\end{equation}
I shall use this choice of penalty operator whenever a specific choice
is necessary in the development of the theory below.

This paper presents a numerically efficient approach to solving the
{\em source subproblem} of problem \ref{eqn:esi}:
\begin{equation}
\label{eqn:esis}
\mbox{given } {\bf c}, \mbox{ choose surface source } \bff \mbox{ to minimize }
\|PL[{\bf c}]^{-1}\bff -d \|^2 + \alpha \|A\bff\|^2 
\end{equation}
Solution of this subproblem is an essential component of {\em variable
  projection} algorithms for solution of the nonlinear inverse problem
\ref{eqn:esi}. Variable projection is not merely a convenient choice
of algorithm for this purpose: it is in some sense essential, see for
example \cite{Symes:SEG20}. It replaces the nonlinear
least squares problem \ref{eqn:esi} with a {\em reduced} problem, to
be solved iteratively. Each iteration involves solution of the
subproblem \ref{eqn:esis}. Therefore efficient solution of the
subproblem is essential to efficient solution of the nonlinear problem
via variable projection.

@@

I will use two 2D numerical models 
throughout to illustrate the theory. In both, horizontal lines serve as
source and receiver surfaces. The first is a ``horizontal crosswell'' or slab
configuration with an acoustic lens positioned between a deep source
and a shallower line of 
receivers, resulting in markedly triplicated arrivals. The goal in
this first example is to construct a surface source that explains the data in a
homogeneous medium (that is, inversion in a wrong velocity, emulating
the early iterations of FWI based on extended modeling). The second is a layered
model in which velocity increases with depth, resulting the formation
of diving waves. This configuration simulates an
ocean-bottom node and a line of near-surface sources: the roles of
source and receiver are switched for computational convenience. In
this second example, the diving wave arrivals are isolated (by an
appropriate mute operator) and a
source constructed that explains them alone.

Since the constructions described here involve a significant number of
components and ideas, I begin the paper with an overview, mapping out
the key steps and their relation to other work.
The next section defines the modeling operator $PL[{\bf c}]^{-1}$ and important specializations (pressure vs. normal velocity
sources and data). It also introduces the two 2D examples mentioned
above.  The following section constructs an approximate inverse of the
modeling operator by time reversal (as suggested by work in
photoacoustic tomography), and illustrates its efficacy via the two
examples. This construction requires extraction of
velocity data, or equivalently a surface source, from pressure data.
The subsequent sections describe this pressure-to-source operator,
express the approximate inverse as the modeling operator adjoint in
weighted norms (thus establishing that the modeling operator is {\em
  approximately unitary} in the sense of these norms), explain how to
use this construction to precondition Conjugate Gradient iteration,
and organize the preconditioning computation so as to involve only one
extra and relatively inexpensive wave propagation calculation. I use
the 2D examples to illustrate each of
these developments. The paper ends with a brief
discussion-and-conclusion section, reviewing what has been
accomplished and listing a few of the many questions left open.

@@

The preconditioner construction is partly inspired by the time
reversal method in photoacoustic tomography
\cite[]{StefanovUhlmannIP:09,Hristova:09}. This problem
clearly has strong similarities to, but also differences from, the
problem studied here.

A simplified mathematical
translation of this medical imaging task is to infer the initial
excess pressure distribution over a fluid-containing region at time
$t=0$ from measurements of the pressure on a surface enclosing the
fluid over a time interval $0 \le t \le t_{\rm max}$. The time reversal method presumes that the
pressure field has returned to equilibrium (zero excess pressure), or
close to it, in the fluid-containing region at the
final time $t=t_{\rm max}$ Then the field can be (at least
approximately) viewed as the backwards-in-time initial boundary value
problem with zero final conditions at $t=t_{\rm max}$, and boundary
values given by the measurements. Evolving the field backwards in time
backwards to $t=0$ thus solves the problem. Except in special
circumstances, the pressure field never actually vanishes at finite
time, so the solution is approximate.

The seismic surface source extension proplem \ref{eqn:esis} differs in
several obvious ways from the photoacoustic setting. The measurement
or receiver surface does not surround the region of wave
propagation. It is not the initial pressure time-slice at $t=0$ that
is to be determined, but a time-extended source $\bff$ confined to a
surface. The penalty operator $A$ has no analogue in the basic
photoacoustic problem description. Finally, once these obstacles are
overcome, using the approximate solution operator to accelerate Krylov
iteration for solution of the optimization problem \ref{eqn:esis}
requires that the operator be identified as the adjoint of the
modeling operator with respect to suitable inner products in its
domain and range. The next few paragraphs sketch resolutions of each
of these issues.

First, reverse-time propagation can be localized via ray theory,
within high-frequency asymptotic approximation. This step requires
some assumptions about the ray fields: all rays responsible for
significant energy in the receiver data must arrive from the source
surface, and must cross the source and receiver surfaces
transversally, that is, making non-zero angles. Also, the rays must
approach source and receiver surfaces from one side or the other
(``the inside''), so that locally the surfaces can be treated as
boundaries of propagation domains. Specifically, rays must emanate
from the source surface towards the inside (``incoming'' rays), and
arrive at the receiver surface from the inside (``outgoing'' rays).
With these assumptions, reverse-time propagation of the receiver data
closely approximates the acoustic fields near the source surface. Next
observe that for a causal field with a surface pressure source (with a
continuous pressure field across the surface), the source is
proportional to the jump in the (particle) velocity field. Moreover,
to leading order in frequency, the velocity field switches sign at the
surface, so the jump is just twice the sampled value on the
surface. Thus reverse-time propagation and reading off the velocity
field on the source surface yields a source that reproduces the
pressure data, within an asymptotically small error.
%Preconditioning amounts to a change of
%norm in the domain and range spaces of the modeling operator. In this
%case, the modfied norms are weighted $L^2$, and the weight operators
%map pressure to corresponding surface source on the source and
%receiver planes.

This is the essence of the approximate solution of the problem
\ref{eqn:esis} for penalty weight $\alpha=0$. To see how Krylov
iteration might be accelerated, note that time-reverse propagation of
the receiver pressure values may be represented by a time-reverse
source propagation, with the source constructed from the velocity
field at the receiver surface. Moreover, the reverse-time
source-to-pressure propagation is the transpose (adjoint) of the
forward-time source-to-pressure propagation - this is a very simple
version of the adjoint state construction \cite[]{Plessix:06}. So the
approximate inverse may be described as: conversion of pressure to source at the
receiver surface, followed by the adjoint of the modeling operator,
followed by the conversion of pressure to source at the source
surface. This sequence precisely describes the adjoint of the modeling
operator with respect to weighted norms in the spaces of source and
receiver data, with the weight operators being the
pressure-to-source operators. That this construction results in an
approximate inverse means that the modeling operator is approximately
unitary with respect to these weighted norms, which in turn implies
that a properly constructed preconditioned conjugate gradient algorithm should converge
rapidly.

@@

For acoustic wave physics, the coefficient vector is
$\bf{c}=(\kappa,\rho)^T$, with components bulk modulus $\kappa$ and
density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of
pressure $p$ (a scalar space-time field) and particle velocity $\bv$
(a vector space-time field). The wave operator $L[\bf{c}]$ is:
\begin{equation}
\label{eqn:aweop}
L[\bf{c}]\bf{u} = 
\left(
\begin{array}{c}
\frac{1}{\kappa}\frac{\partial p}{\partial t}  + \nabla \cdot \bv, \\
\rho\frac{\partial \bv}{\partial t} + \nabla p.
\end{array}
\right) 
\end{equation}
That is,
\begin{equation}
  \label{eqn:awemat}
  L[{\bf c}] = \left(
    \begin{array}{cc}
      \frac{1}{\kappa}\frac{\partial}{\partial t} & \nabla \cdot \\
      \nabla & \rho \frac{\partial}{\partial t}
    \end{array}
  \right)
\end{equation}
$L[{\bf c}]$ has a well-defined inverse %in the sense of distributions
if it is restricted to either causal or anti-causal vector wavefields.

@@

Most of what follows is valid for any space dimension $n >0$. I will
describe the theory for $n=3$, write $\bx=(x,y,z)^T$ for the spatial
coordinate vector, and similarly for the 
particle velocity $\bv = (v_x,v_y,v_z)$. For computational convenience, the
examples are two-dimensional.

The surface source extension replaces point sources on or near a
surface $\Sigma_s$ in $\bR^3$ with source functions confined to the
surface . The surface should be supplied with a choice of unit normal
vector field ${\bf n}_s$. Since the surface separates space into two
non-overlapping parts, at least locally, designate the part into which
the normal vector points as the ``outside'', so that ${\bf n}_s$ is
the outward normal.

For
acoustic modeling, surface sources are combinations of constitutive law
defects and loads normal to the surface, localized on $\Sigma_s$. That
is, right-hand sides in the system $L\bu=\bff$ take the form
$\bff(\bx,t) = (h_s\delta_{\Sigma_s},
f_s{\bf n}_s\delta_{\Sigma_s})^T$ for scalar defect $h_s$ and normal
load $f_s$. With the choice $L$ given in
\ref{eqn:awemat}, the causal system $L\bu=\bff$
takes the form
\begin{eqnarray}
\label{eqn:awepm}
\frac{1}{\kappa}\frac{\partial p}{\partial t} & = & - \nabla \cdot \bv +
h_s \delta_{\Sigma_s}, \nonumber \\
\rho\frac{\partial \bv}{\partial t} & = & - \nabla p +
                                                f_s{\bf n}_s \delta_{\Sigma_s},\nonumber \\
p & =& 0 \mbox{ for }  t \ll 0,\nonumber\\ 
\bv & = & 0 \mbox{ for } t \ll 0.
\end{eqnarray}

\noindent {\bf Remark:} In system \ref{eqn:awepm} and many similar
systems to follow, I will use the shorthand such as
\[
  p = 0 \mbox{ for } t \ll 0 
\]
to mean that $p$ is {\em causal}, that is,
\[
  \mbox{For some } T \in \bR, p(\cdot,t) = 0 \mbox{ for all } t <
  T.
\]
Similarly,
\[
  p = 0 \mbox{ for } t \gg 0 
\]
signifies that $p$ is anti-causal.

Extended forward modeling consists in solving \ref{eqn:awepm} and
sampling the solution components at a receiver surface $\Sigma_r$ with
(``outward'') unit normal field ${\bf n}_r$.
$P_s$ and $P_r$ are the sampling operators on $\Sigma_s$ and
$\Sigma_r$ respectively. In
practice, sampling necessarily occurs at a discrete array of points
(trace locations), and over a zone of finite extent. In this
theoretical discussion, I will neglect the finite sample rate, and
regard the data, for example $P_rp$, as continuously sampled. The
output samples are necessarily muted, that is, non-zero only over a space-time domain
of finite extent. This mute, and any tapering applied to the data
traces, are regarded as part of the sampling operators $P_s,P_r$.

The vector modeling operator ${\cal S}$ relates the source amplitudes
$h_s,f_s$ on $\Sigma_s$ to the sampled pressure and normal velocity
$p$, $\bv \cdot {\bf n}_r$ for 
the solution $(p,\bv)$ of the systems \ref{eqn:awepm} by
\begin{equation}
  {\cal S}(h_s,f_s)^T  = (P_rp,P_r(\bv \cdot {\bf n}_r) )^T,
  \label{eqn:fwd}
\end{equation}

Denote by $\Pi_i, i=0,1$ the projection on the first,respectively
second, component of a vector in $\bR^2$. The forward modeling
operator from pressure source to pressure trace is
\begin{equation}
  \label{eqn:sdef}
  S = \Pi_0 {\cal S}\Pi_0^T 
\end{equation}
%and the forward modeling operator from velocity source (normal force)
%to velocity trace is
%\begin{equation}
%  \label{eqn:vdef}
%  V_{z_s,z_r} = \Pi_1 {\cal S}_{z_s,z_r} \Pi_1^T 
%\end{equation}

With these conventions, we can write the version of the source
subproblem \ref{eqn:esis} studied in this paper as: given a pressure
gather $d$ on $\Sigma_r$,
\begin{equation}
  \label{eqn:esisp}
  \mbox{find }h_s\mbox{ to minimize }\|Sh_s- d\|^2 +
  \alpha^2\|Ah_s\|^2.
\end{equation}

The two examples mentioned in the introduction illustrate the setting
just described. The first example (Figure \ref{fig:olensrays0}) embeds
an acoustic (low-velocity) lens between a source surface
$\Sigma_s= \{z = z_s=3000 \mbox{ m}\}$ (orange horizontal line) m and
a receiver surface $\Sigma_r= \{z=z_r=1000 \mbox{ m}\}$ (yellow
horizontal line).  The data to be used in this example results from a
point source at $(3500, 3500)$ m (orange star). I have overlain
the rays connecting this source point with the data portion of the
receiver surface: evidently all of the rays involved cross the source
and receiver surfaces transversally. Also, regarding ``inwards'' as
being ``upwards'' at the source surface, ``downwards'' at the receiver
surface, it is clear that the data is incoming at the source surface,
outgoing at the receiver surface.

@@

%\ref{fig:drecplh0}, \ref{fig:dfwdplh0}
Figure \ref{fig:drecplh0} shows the gather $P_rp_{\rm pt}$, where
$p_{\rm pt}$ is causal field generated by the point source at
$(3500,3500)$ m with a trapezoidal bandpass filter wavelet having
significant energy between 1 and 12.5 Hz. The mute embedded in $P_r$
limits trace positions to 2000 m $ \le x_r \le $ 6000 m, and times to
1.2 s $ \le t \le $ 3 s.

Figure \ref{fig:dsrcphh0} displays an extended (pressure) source $h_s$, in the form of
traces in the spatial range 2000 m $ \le x_r \le $ 6000 m and time
range 0 s $\le t \le $ 2 s. The modeling operator output $Sh_s$,
obtained from the causal solution $(p, \bv)$ of \ref{eqn:awepm}
with this choice of $h_s$ and $f_s=0,$ by application of the sampling
operator $P_r$ to $p$, is shown in Figure \ref{fig:dfwdplh0}.

\noindent {\bf Remark:} The close resemblance between Figures
\ref{fig:drecplh0} and \ref{fig:dfwdplh0} is not accidental  - the
extended source $h_ s$ shown in Figure \ref{fig:dsrcphh0} is constructed
so that $Sh_s$ (Figure \ref{fig:dfwdplh0}) closely
approximates the point source gather $P_r p_{\rm pt}$ (Figure
\ref{fig:drecplh0}). This construction will be explained in the next section.

@@

This geometry permits inversion of the diving wave data alone, which
is isolated via a mute, depicted in Figure \ref{fig:ndwmute0}. The
mute includes time truncation at 6 s, and offsets between -10000 and
10000 m. The isolated
diving wave (window between offsets 4000 and 10000) is shown in
Figure \ref{fig:ndwdata0}.

\begin{figure}
  \centering
  \subfloat[\label{fig:ndwraw0}]{\includegraphics[width=0.32\textwidth]{ndwraw0.pdf}}
  \subfloat[\label{fig:ndwmute0}]{\includegraphics[width=0.32\textwidth]{ndwmute0.pdf}}
  \subfloat[\label{fig:ndwdata0}]{\includegraphics[width=0.32\textwidth]{ndwdata0.pdf}}
  \caption{Data generated using configuration in Figure
    \ref{fig:ooplrays0}. (a) traces from point source at $(500,10000)$ m
    with $[1, 2, 7.5, 12.5]$ Hz zero phase trapezoidal bandpass
    filter wavelet, delayed by $0.4$ s. (b) Mute function,
    incorporated in the sampling operator $P_r$. (c) Diving wave
    gather. Color scale is same for (a) and (c).}
\end{figure}
